---
title: "Cyclistic"
author: "Toyin Olapade"
date: "11/28/2021"
output: pdf_document
---

```{r}
library(tidyverse)
library(lubridate)
```


```{r}
dt_202004 <- read.csv("202004-divvy-tripdata.csv")
dt_202005 <- read.csv("202005-divvy-tripdata.csv")
dt_202006 <- read.csv("202006-divvy-tripdata.csv")
dt_202007 <- read.csv("202007-divvy-tripdata.csv")
dt_202008 <- read.csv("202008-divvy-tripdata.csv")
dt_202009 <- read.csv("202009-divvy-tripdata.csv")
dt_202010 <- read.csv("202010-divvy-tripdata.csv")
dt_202011 <- read.csv("202011-divvy-tripdata.csv")
dt_202012 <- read.csv("202012-divvy-tripdata.csv")
dt_202101 <- read.csv("202101-divvy-tripdata.csv")
dt_202102 <- read.csv("202102-divvy-tripdata.csv")
dt_202103 <- read.csv("202103-divvy-tripdata.csv")
dt_202104 <- read.csv("202104-divvy-tripdata.csv")
dt_202105 <- read.csv("202105-divvy-tripdata.csv")
dt_202106 <- read.csv("202106-divvy-tripdata.csv")
dt_202107 <- read.csv("202107-divvy-tripdata.csv")
dt_202108 <- read.csv("202108-divvy-tripdata.csv")
dt_202109 <- read.csv("202109-divvy-tripdata.csv")
dt_202110 <- read.csv("202110-divvy-tripdata.csv")

```
```{r}
dt <- rbind(
  dt_202004,
  dt_202005,
  dt_202006,
  dt_202007,
  dt_202008,
  dt_202009,
  dt_202010,
  dt_202011,
  dt_202012,
  dt_202101,
  dt_202102,
  dt_202103,
  dt_202104,
  dt_202105,
  dt_202106,
  dt_202107,
  dt_202108,
  dt_202109,
  dt_202110
)
```

```{r}
summary(dt)
head(dt)
```
While ha;ndinek



```{r}
sum(is.na(dt)) # 199,067

dt %>%
  filter(start_station_name == "" | end_station_name == "") %>%
  count() # 982,947
```

```{r}
head(dt)
```

```{r}
# separate(started_at, into = date("yyyy/mm/dd", "started_time"),sep = ( ) )

dt_202102 %>%
  date(started_at) %>%
  head(3)

ymd(date("2020-04-26 17:45:14"))
```


* What are the most frequently used ride type?
* When do riders ride the most?
* How does ride duration differ by ride type?
* Which member type ride with us the most?


* What are the most frequently used ride type?

```{r}
dt %>%  
  group_by(rideable_type) %>%
  count()

  
```


* When do riders ride the most?
* How does ride duration differ by ride type?
* Which member type ride with us the most?



Three questions will guide the future marketing program:
1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?


1. How do annual members and casual riders use Cyclistic bikes differently?

```{r}
dt %>%
  group_by(member_casual) %>% count()

dt %>%
  group_by(member_casual, rideable_type) %>% 
  count() 

dt %>%
  group_by(member_casual, rideable_type) %>% 
  summarize(
    n=n()
  ) %>%
  mutate(
    pct_ride <- paste0(round(100 * n/sum(n), 1), "%")
  ) 
```


2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?












